---
title: "Week 6 Assignment"
author: "Rajesh Adhi"
date: "2025-03-03"
output: html_document
---
## Loading all the libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(sf)
library(geosphere)
library(dplyr)
library(GA)
```

# Problem Statement
Identifying 8 FQHCs in Missouri to introduce mental health services, prioritizing locations with the highest residential density within a 30-mile radius.

# Data Loading and Preparation

```{r load-data}
# Load the population data
pop.data <- data.table::fread("Mo_pop_Sim\\Mo_pop_Sim.csv")
frac <- 0.0001
small.data <- pop.data[sample(1:nrow(pop.data),
                             size = round(nrow(pop.data) * frac),
                             replace = F)]

# Load the FQHC data
data_path <- 'MO_2018_Federally_Qualified_Health_Center_Locations\\MO_2018_Federally_Qualified_Health_Center_Locations\\MO_2018_Federally_Qualified_Health_Center_Locations.shp'
fqhc.data <- data.table(as.data.frame(st_read(data_path,
                   'MO_2018_Federally_Qualified_Health_Center_Locations')))

fqhc.data <- fqhc.data[, .(OBJECTID, Facility, City, County, Longitude, Latitude)]
small.data <- small.data[, .(long, lat)]
cat("Number of sampled population points:", nrow(small.data), "\n")
cat("Number of FQHCs:", nrow(fqhc.data), "\n")
```

# Simple Coverage Analysis

```{r}
# Function to count residents within a 30-mile radius
count_residents <- function(fqhc_idx) {
  fqhc_loc <- fqhc.data[fqhc_idx]
  distances <- distHaversine(
    matrix(c(fqhc_loc$Longitude, fqhc_loc$Latitude), ncol = 2),
    matrix(c(small.data$long, small.data$lat), ncol = 2)
  ) / 1609.34  # Convert meters to miles
  count <- sum(distances <= 30)
  return(count)
}
```

# Calculating thee coverage for each FQHC and displaying top 8 by simple coverage
```{r}
coverage_counts <- sapply(1:nrow(fqhc.data), count_residents)
fqhc.data$coverage <- coverage_counts
fqhc.data$estimated_total <- coverage_counts * (1/frac)
fqhc.data <- fqhc.data[order(-coverage)]

cat("Top 8 FQHCs by population count in 30-mile radius:\n")
top_simple <- fqhc.data[1:8]
print(top_simple[, .(OBJECTID, Facility, City, County, coverage, estimated_total)])
```

# GA Implementation
# Function to calculate the fitness of a set of FQHCs We want to maximize total unique population coverage
```{r}
fitness_function <- function(selected_indices) {
  covered_residents <- rep(FALSE, nrow(small.data))
  for (idx in selected_indices) {
    fqhc_loc <- fqhc.data[idx]
    distances <- distHaversine(
      matrix(c(fqhc_loc$Longitude, fqhc_loc$Latitude), ncol = 2),
      matrix(c(small.data$long, small.data$lat), ncol = 2)
    ) / 1609.34  # Converting meters to miles
    # Marking residents as covered if within 30 miles
    covered_residents <- covered_residents | (distances <= 30)
  }
  return(sum(covered_residents))
}
cat("Running genetic algorithm to optimize FQHC selection...\n")
ga_result <- ga(
  type = "permutation",
  fitness = function(indices) fitness_function(indices[1:8]),  # Only use first 8 indices
  lower = 1,
  upper = nrow(fqhc.data),
  popSize = 30,
  maxiter = 15,
  run = 5,
  pmutation = 0.2,
  monitor = FALSE
)
```

```{r}
best_indices <- ga_result@solution[1, 1:8]
selected_fqhcs <- fqhc.data[best_indices]

# Calculate total unique coverage
unique_coverage <- fitness_function(best_indices)
estimated_total <- unique_coverage * (1/frac)

cat("\nGA-optimized selection of 8 FQHCs:\n")
print(selected_fqhcs[, .(OBJECTID, Facility, City, County)])

cat("\nTotal unique population coverage (sample):", unique_coverage, "\n")
cat("Estimated total unique population coverage:", round(estimated_total), "\n")
```

# Individual Coverage Analysis
```{r individual-coverage}
# Calculate individual coverage for selected FQHCs
individual_coverage <- sapply(best_indices, count_residents)
selected_fqhcs$coverage <- individual_coverage
selected_fqhcs$estimated_total <- individual_coverage * (1/frac)

cat("Coverage by each selected FQHC:\n")
print(selected_fqhcs[, .(Facility, City, coverage, estimated_total)])
```

# Conclusion

I've chosen 8 FQHC's to introduce mental health services, focusing on the areas where the most people live within a 30-mile radius. Using a genetic algorithm, we found the best way to cover the most unique individuals with these services.

**Why This Approach?**

1. It focuses on where the most could be patients are, making it easier for them to access mental health care.
2. The GA helps minimize overlap between different service areas, ensuring that more people are served.
3. A 30-mile distance is practical for people traveling to get mental health support.

**What We Assumed**

1. Everyone has a similar need for mental health services, no matter their background.
2. Each FQHC is equally capable of offering these new services.
3. A 30-mile trip is a reasonable distance for people to travel for help.
4. Everyone in that radius can access the services equally.

This process allowed us to pick FQHCs that together reach more unique residents than just selecting the 8 with the highest individual coverage.
